---
title: "Practical Machine Learning Peer Review Project"
author: "Matt Zapf"
date: "October 21, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Coursera Practical Machine Learning Project:Prediction Assignment Writeup


##I. Overview
```
This HTMl document is the report of the Peer Assessment project from Coursera's course Practical Machine Learning, as part of the Specialization in Data Science. 
The analysis to be the basis for the course quiz and a prediction assignment writeup. The main goal of the project is to predict the manner in which 6 participants performed the "classe" variable in the training set. 
```


##II. Background
```
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).
```


##III. Data Loading and Exploratory Analysis


###1) Dataset Overview
```
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from http://groupware.les.inf.puc-rio.br/har. Full source:

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. "Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13)". Stuttgart, Germany: ACM SIGCHI, 2013.

My special thanks to the above mentioned authors for being so generous in allowing their data to be used for this kind of assignment.

A short description of the datasets content from the authors' website:

"Six young health participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different fashions: exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E).

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes. Participants were supervised by an experienced weight lifter to make sure the execution complied to the manner they were supposed to simulate. The exercises were performed by six male participants aged between 20-28 years, with little weight lifting experience. We made sure that all participants could easily simulate the mistakes in a safe and controlled manner by using a relatively light dumbbell (1.25kg)."
```

###2) Environment Preparation
```
First upload the R libraries that are necessary for the complete analysis.
```
```{r echo=TRUE}
rm(list=ls())                # free up memory for the download of the data sets
#setwd("C:\Users\mattz\Desktop\Coursera")
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
set.seed(123)
```

###3) Data Loading and Cleaning
```
The next step is loading the dataset from the URL provided. The training dataset is then partinioned into two pieces to create a training set (70% of the data) for the modeling process and a Test set (with the remaining 30%) for the validations. The testing dataset is not changed and will only be used for the quiz results generation.
```

```{r echo=TRUE}
# set the URL for the download
UrlTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
UrlTest  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

# download the datasets
training <- read.csv(url(UrlTrain))
testing  <- read.csv(url(UrlTest))

# create a partition with the training dataset 
inTrain  <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
TestSet  <- training[-inTrain, ]
dim(TrainSet)
```

```{r echo=TRUE}
dim(TestSet)
```

```
Both of the created datasets have 160 variables. Those variables have contain values NA. These can be removed with the cleaning process below. The Near Zero variance (NZV) variables are also removed and the ID variables as well.
```

```{r echo =TRUE}
# remove variables with Nearly Zero Variance
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
TestSet  <- TestSet[, -NZV]
dim(TrainSet)
```

```{r echo =TRUE}
dim(TestSet)
```

```{r echo =TRUE}
# remove variables that are mostly NA
SetNA    <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95
TrainSet <- TrainSet[,SetNA==FALSE]
TestSet  <- TestSet[, SetNA==FALSE]
dim(TrainSet)
```

```{r echo =TRUE}
dim(TestSet)
```

```{r echo =TRUE}
# remove identification only variables (columns 1 to 5)
TrainSet <- TrainSet[, -(1:5)]
TestSet  <- TestSet[, -(1:5)]
dim(TrainSet)
```

```{r echo =TRUE}
dim(TestSet)
```
```
After the cleaning process above, the number of variables for the analysis has been reduced to 54.
```
###4) Correlation Analysis

```
Correlation among variables is analyzed next before proceeding to the modeling procedures.
```

```{r echo =TRUE}
corMat<- cor(TrainSet[, -54])
corrplot(corMat, order = "FPC", method = "color", type = "lower", 
         tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```

```
The highly correlated variables are shown in dark colors in the graph. We could also do a Principal Compenents Analysis as processing step to review the data.  However, as the correlations are quite few, this step will be skipped for this assignment.
```

##IV. Prediction Model Building
```
Three methods will be applied to analyze the regressions in the Train dataset and the one with higher accuracy when applied to the Test dataset will be used for the quiz predictions. The methods are: Random Forests, Decision Tree and Generalized Boosted Model.
A Confusion Matrix is plotted at the end of each analysis to better visualize the accuracy of the models.
```

###1) Method: Random Forest
```{r echo =TRUE}
# model fit
set.seed(123)
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modelFitRandForest <- train(classe ~ ., data=TrainSet, method="rf",
                          trControl=controlRF)
modelFitRandForest$finalModel

```

```{r echo =TRUE}
# prediction on Test dataset
predictRandForest <- predict(modelFitRandForest, newdata=TestSet)
confMatrixRandForest <- confusionMatrix(predictRandForest, TestSet$classe)
confMatrixRandForest
```
```{r echo =TRUE}
# plot matrix results
plot(confMatrixRandForest$table, col = confMatrixRandForest$byClass, 
     main = paste("Random Forest - Accuracy =",
                  round(confMatrixRandForest$overall['Accuracy'], 4)))
```
###2) Method: Decision Trees

```{r echo =TRUE}
# model fit
set.seed(123)
modelFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class")
fancyRpartPlot(modelFitDecTree)

```
```{r echo =TRUE}
# prediction on Test dataset
predictDecTree <- predict(modelFitDecTree, newdata=TestSet, type="class")
confMatrixDecTree <- confusionMatrix(predictDecTree, TestSet$classe)
confMatrixDecTree
```

```{r echo =TRUE}
# plot matrix results
plot(confMatrixDecTree$table, col = confMatrixDecTree$byClass, 
     main = paste("Decision Tree - Accuracy =",
                  round(confMatrixDecTree$overall['Accuracy'], 4)))
```
###3) Method: Generalized Boosted Model

```{r echo =TRUE}
# model fit
set.seed(123)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modelFitGBM  <- train(classe ~ ., data=TrainSet, method = "gbm",
                    trControl = controlGBM, verbose = FALSE)
modelFitGBM$finalModel
```

```{r echo =TRUE}
# prediction on Test dataset
predictGBM <- predict(modelFitGBM, newdata=TestSet)
confMatrixGBM <- confusionMatrix(predictGBM, TestSet$classe)
confMatrixGBM
```

```{r echo =TRUE}
# plot matrix results
plot(confMatrixGBM$table, col = confMatrixGBM$byClass, 
     main = paste("GBM - Accuracy =", round(confMatrixGBM$overall['Accuracy'], 4)))
```

##V. Applying the Selected Model to the Test Data

```
The accuracy of the 3 regression models are:

Random Forest : 0.9963
Decision Tree : 0.7368
GBM : 0.9839
With these results, the Random Forest model will be applied to predict the 20 quiz results, testing dataset as shown below.
```
```{r echo =TRUE}
predictTEST <- predict(modelFitRandForest, newdata=testing)
predictTEST
```